---
title: "GuestLecture_ANOVA_Markdown"
author: "Jessica Macaluso"
date: "2025-02-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Storytelling: The Fundamentals of Data Visualization with ggplot2


We will be exploring the education data set that was retrieved from kaggle.com (https://www.kaggle.com/datasets/desalegngeb/students-exam-scores?resource=download).


## 1. Load Packages

```{r, packages}
library(tidyverse)
library(visdat)
library(corrplot)
```

## 2. Read in Data

```{r, data}
education <- read.csv("education.csv")
head(education)
```

## 3. Overview of Data

```{r, overview}
nrow(education)
ncol(education)
education %>% map_dbl(~sum( is.na(.) ) ) %>% sum()
```

30641 rows, 16 columns, and 13901 missing values.

## 4. Missingness

```{r, missingness}
vis_dat(education)
```

## 5. Data Dictionary
- Gender: Gender of the student (2 categories: male or female)
- EthnicGroup: Ethnic group of the student (5 categories: group A, group B, group C, group D, group E)
- ParentEduc: Parent(s) education background (6 categories: some high school, high school, some college, associate’s degree, bachelor’s degree, master’s degree) 
- LunchType: School lunch type (2 categories: standard or free/reduced)
- TestPrep: Test preparation course followed (2 categories: none or completed)
- ParentMaritalStatus: Parent(s) marital status (4 categories: single, married, widowed, divorced)
- PracticeSport: How often the student practice sport (3 categories: sometimes, regularly, never)
- IsFirstChild: If the child is first child in the family or not (2 categories:
yes or no)
- NrSiblings: Number of siblings the student has (8 categories: 0, 1, 2, 3, 4, 5, 6, 7)
- TransportMeans: Means of transport to school (2 categories: school bus or private)
- WklyStudyHours: Weekly self-study hours (3 categories: less that 5 hours, between 5 and 10 hours, or more than 10 hours)
- MathScore: Math test score (0-100)
- ReadingScore: Reading test score (0-100)
- WritingScore: Writing test score (0-100)
- School: 31 schools (31 categories: 1-31)

## 6. Change Datatypes

```{r, datatypes}
education %>% mutate(student_ID = as.factor(student_ID),
                     Gender = as.factor(Gender),
                     EthnicGroup = as.factor(EthnicGroup),
                     ParentEduc = as.factor(ParentEduc),
                     LunchType = as.factor(LunchType),
                     TestPrep = as.factor(TestPrep),
                     ParentMaritalStatus = as.factor(ParentMaritalStatus),
                     PracticeSport = as.factor(PracticeSport),
                     IsFirstChild = as.factor(IsFirstChild),
                     NrSiblings = as.factor(NrSiblings),
                     TransportMeans = as.factor(TransportMeans),
                     WklyStudyHours = as.factor(WklyStudyHours),
                     MathScore = as.numeric(MathScore),
                     ReadingScore = as.numeric(ReadingScore),
                     WritingScore = as.numeric(WritingScore),
                     School = as.factor(School)) -> education
glimpse(education)
```

Now all categorical variables are factors and all continuous variables are numeric.

## 7. Categorical Variables (Simple Bar Chart)

```{r, simple_barchart}
education %>% count(Gender) %>%
  ggplot(mapping = aes(x = Gender, y = n)) +
  geom_col(fill = 'pink') +
  theme_bw()
```

## 8. Categorical Variables (Stacked Bar Chart)

```{r, stacked_barchart}
education %>%
  ggplot(mapping = aes(y = ParentEduc)) +
  geom_bar(mapping = aes(fill = LunchType)) +
  theme_classic()
```

## 9. Categorical Variables (Dodged Bar Chart)

```{r, dodged_barchart}
education %>%
  ggplot(mapping = aes(y = ParentEduc)) +
  geom_bar(mapping = aes(fill = LunchType), position = 'dodge') +
  theme_classic()
```

## 10. Categorical Variables (Heat Map)

```{r, heatmap}
education %>%
  count(ParentEduc, LunchType, Gender) %>%
  ggplot (mapping = aes(x = LunchType, y = ParentEduc)) +
  geom_tile(mapping = aes(fill = n),
            color = "black") + 
  facet_wrap(~ Gender, labeller = 'label_both') +
  geom_text(mapping = aes(label = n), color = 'white') + 
  theme_classic()
```

## 11. Continuous Variables (Histogram)

```{r, histogram}
education %>%
  ggplot(mapping = aes(x = ReadingScore)) +
  geom_histogram(bins = 30, fill = 'green') +
  theme_classic()
```

## 12. Continuous Variables (Density Plot)

```{r, densityplot}
education %>%
  ggplot(mapping = aes(x = MathScore)) +
  geom_density() +
  theme_bw()
```

## 13. Continuous Variables (Histogram + Density Plot)

```{r, histogram_densityplot}
education %>%
  ggplot(mapping = aes(x = MathScore)) +
  geom_histogram(bins = 30, fill = 'grey', mapping = aes(y = after_stat(density))) + 
  geom_density(color = 'purple', linewidth = 1.25) + 
  theme_minimal()
```

## 14. Continuous Variables (Scatterplot)

```{r, scatterplot}
education %>%
  ggplot(mapping = aes(x = WritingScore, y = ReadingScore)) +
  geom_point(size = 2.5 , alpha = .25) +
  theme_bw()
```

## 15. Continuous Variables (Correlation Plot)

```{r, correlationplot}
education %>%
  select(MathScore, ReadingScore, WritingScore) %>%
  cor() %>% 
  corrplot::corrplot(method = 'number', type = 'upper')
```

## 16. Categorical + Continuous Variables (Boxplot)

```{r, boxplot}
education %>%
  ggplot(mapping = aes(x = WklyStudyHours, y = MathScore)) +
  geom_boxplot()
```

## 17. Categorical + Continuous Variables (Violin Plot)

```{r, violinplot}
education %>%
  ggplot(mapping = aes(x = WklyStudyHours, y = MathScore)) + 
  geom_violin(fill = 'grey') + 
  theme_bw()
```

## 18. Categorical + Continuous Variables (Scatterplot)

```{r, scatterplot2}
education %>%
  ggplot(mapping = aes(x = ReadingScore, y = MathScore)) + 
  geom_point(mapping = aes(color = Gender, shape = Gender), size = 3) +
  ggthemes::scale_color_colorblind() + 
  theme_bw()
```

## 19. Categorical + Continuous Variables (Heat Map)

```{r, heatmap2}
education %>%
  ggplot(mapping = aes(x = ReadingScore, y = MathScore)) +
  geom_bin_2d(bins = 11) + 
  facet_wrap(~ ParentEduc) +
  scale_fill_viridis_c() + 
  theme_bw()
```
