---
title: "R Basics"
author: "Jessica Macaluso"
date: "2025-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

```{r cars}
library(psych)
library(tidyverse)
library(nycflights13)
```

## Look at Data

```{r data1}
glimpse(flights)
```

Let's look at the first six rows of the data.

```{r data2}
head(flights)
```

Let's look at the first twenty rows of the data.

```{r data3}
head(flights, 20)
```

Let's look at the different variable types (aka the"kind" of variable: character, factor, numeric, integer, etc.).

```{r class}
class(flights) 
```

## Using `filter()`

`select()` all variables except carrier THEN `filter()` to only include observations whose destination is ORD.

```{r filter1}
flights %>%
  select(-carrier) %>%
  filter(dest == "ORD")
```

`filter()` all flights on January 1st.

```{r filter2}
flights %>% filter(month == 1, day == 1) 
```

`filter()` all flights on January 1st and save into a new data frame.

```{r filter3}
jan1 <- flights %>% filter(month == 1, day == 1)
```

`filter()` all flights on December 25th, save into a new data frame, and `print()`.

```{r filter4}
(dec25 <- flights %>% filter(month == 12, day == 25))
```

`filter()` all flights on December 25th, save into a new data frame, and `print()`.

```{r filter5}
(dec25 <- flights %>% filter(month == 12, day == 25))
```
 
`filter()` all flights in either November or December.

```{r filter6}
flights %>% filter(month == 11 | month == 12)
```

`filter()` all flights not delayed by two or more hours.

```{r filter7}
flights %>% filter(!(arr_delay > 120 | dep_delay > 120))
```

`filter()` all flights that departed in summer (July, August, and September).

```{r filter8}
flights %>% filter(month >= 7 & month <= 9)
```

`filter()` all flights that flew to Houston (IAH or HOU).

```{r filter9}
flights %>% filter((dest == 'IAH' | dest == 'HOU')) 
```

## Using `arrange()`

`arrange()` flights by descending order.

```{r arrange}
flights %>% arrange(desc(dep_delay)) 
```

## Using `select()`

`select()` year, month, and day.

```{r select1}
flights %>% select(year, month, day)
```

`select()` select all columns between year and day (inclusive).

```{r select2}
flights %>% select(year:day)
```

`select()` select all columns except those from year to day (inclusive).

```{r select3}
flights %>% select(year:day)
```
 
## Using `summarise()`

`summarise()` by removing NAs prior to computation.

```{r summarise}
flights %>% summarise(delay = mean(dep_delay, na.rm = TRUE))
``` 
 
## Using `mutate()`

Use `mutate()` to create new variables.

```{r mutate1}
flights %>% mutate(gain = dep_delay - arr_delay, hours = air_time / 60, 
         gain_per_hour = gain / hours)
``` 
 
```{r mutate2}
flights %>%
  mutate(gain = dep_delay - arr_delay, 
         hours = air_time / 60, 
         gain_per_hour = gain / hours, 
         fast_plane = ifelse(gain_per_hour > 1, 1, 0))
``` 

## Descriptive Statistics
This is an introduction to `dplyr` for Faster Data Manipulation in R. There are five basic verbs: `filter()`, `select()`, `arrange()`, `summarise()` (plus `group_by()`), and`mutate()`
- Pick observations by their values (`filter()`); 
- Reorder the rows (`arrange()`);
- Pick variables by their names (`select()`);
- Create new variables with functions of existing variables (`mutate()`);
- Collapse many values down to a single summary (`summarise()`)

These can all be used in conjunction with `group_by()` which changes the scope of each function from operating on the entire dataset to operating on it group-by-group.

```{r describe}
describe(flights$dep_time)
``` 

```{r mean}
mean(flights$dep_time, na.rm = TRUE)
``` 

```{r median}
median(flights$dep_time, na.rm = TRUE)
``` 

```{r standard_deviation}
sd(flights$dep_time, na.rm = TRUE)
``` 
