---
title: "Regression"
author: "Jessica Macaluso"
date: "2025-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

```{r packages}
library(MASS)
library(dplyr)
library(ggplot2)
library(psych)
library(ppcor)
```

## Disable Scientific Notation

```{r scipen}
options(scipen = 999)
```

A researcher is interested in examining whether reading skills (READING) at kindergarten entry is predicted by cognitive stimulation (COGSTIM) in the home environment and is concerned about the confounding influence of children’s age when they start kindergarten (AGEENTRY).

## Disable Scientific Notation

```{r scipen_1}
options(scipen = 999)
```

#### Read in data

```{r data1}
data_1 <- read.csv("regression_df.csv") 
```

```{r data2}
head(data_1)
```
Variables:
- READING = reading skills (DV)
- COGSTIM = cognitive stimulation (IV)
- AGEENTRY = are when children start kindergarten (IV)

Let's calculate the regression equation predicting reading skills with cognitive stimulation and age at kindergarten entry. Present the standardized and unstandardized regression equations.

## Unstandardized

```{r model_1a}
model_1 <-  lm(READING ~ COGSTIM + AGEENTRY, data= data_1)
summary(model_1)
```

```{r model_1b}
anova(model_1)
```

```{r model_1c}
confint(model_1,level=0.95)
```

```{r model_1d}
coefficients(model_1)
```

The intercept is -3.59 and is the predicted level of reading skills (READING) when cognitive stimulation (COGSTIM) and age at kindergarten entry (AGEENTRY) are zero. A one unit increase in reading skills (READING) is associated with a 6.73 unit increase in cognitive stimulation (COGSTIM) and a 0.29 unit increase in age at kindergarten entry (AGEENTRY).

Y = -3.59 + 6.73X1 + 0.29X2

For COGSTIM, t-value = 35.10 with a significant p-value of 0.0000000000000002. For AGEENTRY, t-value = 14.81 with a significant p-value of 0.0000000000000002. 

## Standardized

```{r standardized}
data_1 <- data_1 %>%
  mutate(zREADING = scale(READING))
data_1 <- data_1 %>%
  mutate(zCOGSTIM = scale(COGSTIM))
data_1 <- data_1 %>%
  mutate(zAGEENTRY = scale(AGEENTRY))
head(data_1)
```

```{r model_1d1}
model_1_z <- lm(zREADING ~ zCOGSTIM + zAGEENTRY, data= data_1) 
summary(model_1_z)
```
 
```{r model_1d2}
anova(model_1_z)
```

```{r model_1d3}
confint(model_1_z, level=0.95) 
```

```{r model_1d4}
coefficients(model_1_z) 
```

The intercept is 4.961146e-16 and is the predicted level of reading skills when cognitive stimulation (COGSTIM) and age at kindergarten entry (AGEENTRY) are zero. A one standard deviation unit increase in reading skills is associated with a 0.40 unit increase in cognitive stimulation and a 0.17 unit increase in age at kindergarten entry.

Y = 0.40X1 + 0.17X2

For COGSTIM, t-value = 35.10 with a significant p-value of 0.0000000000000002. For AGEENTRY, t-value = 14.81 with a significant p-value of 0.0000000000000002.

What proportion of the total variance in reading skills is explained by cognitive stimulation AND age at kindergarten entry? Adjusted R^2 so either is fine.

m1 <- lm(READING ~ COGSTIM + AGEENTRY, data= data_1) 
summary(m1)
m1_r<-summary(m1) 
m1_r$r.squared
m1_r$r.squared - summary(lm(READING~ READING, data=data_1))$r.squared
# 0.18 (18%) represents the total variance in reading skills (READING) that is 
  # explained by both cognitive stimulation (COGSTIM) and age at kindergarten
  # entry (AGEENTRY).

# Is the proportion in c significantly different from zero?
resid(model_1) 
model_1.res <- resid(model_1) 
plot(data_1$READING, model_1.res)

```{r m1a}
m1 <- lm(READING ~ COGSTIM + AGEENTRY, data= data_1) 
summary(m1)
```

```{r m1b}
m1_r<-summary(m1) 
m1_r$r.squared
```

```{r m1c}
m1_r$r.squared - summary(lm(READING~ READING, data=data_1))$r.squared 
```

R^2 = variance explained by the model/total variance

Yes 0.18 is significantly different from zero because our F statistic is above zero [F(2,6396)=720.7] and our p-value is significant (p=0.00000000000000022).
