---
title: "Two Way ANOVA"
author: "Jessica Macaluso"
date: "2025-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load in packages

```{r, load packages}
library(tidyverse)
library(DescTools)
library(effectsize)
library(emmeans)
library(car)
library(reghelper)
```
## Disable scientific notation

```{r, disable scientific notation}
options(scipen = 999) 
```


## VARIABLES & STUDY PREMISE
- 120 participants, 40/group, between-subjects design
- DV - performance on final retrieval quiz
- VariableA (IV/PV) - level of trained expertise (categorical, binomial)
  - A1 - Novice (lower mean)
  - A2 - Expert (higher mean)
- VariableB (IV/PV) - study strategy intervention (categorical, binomial)
  - B1 - Restudy Intervention (lower mean)
  - B2 - Retrieval Intervention (higher mean)

## Read in the data
```{r, data1}
final <- read.csv("finalexam-2way.csv")
head (final)
```

Treat as factors:

```{r, data2}
final <- read.csv('finalexam-2way.csv', stringsAsFactors = TRUE)
head(final)
```

## One-way ANOVA VariableA

```{r, model_a}
model_a <- lm(DV~ 1 + VariableA, data=final)
anova(model_a)
```

No, there is not significant difference between groups. Same performance regardless of assigned expertise.

## One-way ANOVA VariableB

```{r, model_b}
model_b <- lm(DV~ 1 + VariableB, data=final)
anova(model_b)
```

Yes, there is a significant difference between groups. Different performance regarding study strategy intervention.

```{r, model_1}
model_1 <- lm(DV ~ VariableA*VariableB , data=final)
summary(model_1)
```

## Visualizing the data

```{r, graph}
final %>% group_by(VariableA, VariableB) %>% 
  summarize(DV_M=mean(DV)) %>%
  ggplot(aes(x=VariableB, y=DV_M, color=VariableA, group=VariableA)) +   
  geom_point() + geom_line()
```

Performance on final quiz (z-scores): A2B2 best (Expert, Retrieval); A1B2 second (Novice, Retrieval); A1B1 third (Novice, Restudy); A2B1 last (Expert, Restudy).

## Contrasts

```{r, contrasts1}
contrasts(final$VariableA)
```

```{r, contrasts2}
contrasts(final$VariableA) <- cbind(c(.5, -.5))
colnames(contrasts(final$VariableA)) <- c('Restudy versus Retrieval on Expertise')
contrasts(final$VariableA)
```

```{r, model_x}
model_x <- lm(DV~ 1 + VariableA, data=final)
summary(model_x)
```

```{r, contrasts3}
contrasts(final$VariableB)
```

```{r, contrasts4}
contrasts(final$VariableB) <- cbind(c(.5, -.5))
colnames(contrasts(final$VariableB)) <- c('Novice versus Expert on Study Strategy')
contrasts(final$VariableB)
```

```{r, model_y}
model_y <- lm(DV~ 1 + VariableB, data=final)
summary(model_y)
```

## Tukey Test

```{r, tukey}
TukeyHSD(aov(model_1))
```


